---
title: "Reproducible Research Assignment Week 2"
author: "WW van Niekerk"
date: "10 March 2017"
output: html_document
---

# Course Assignment: Week 2

## Step 1: Getting the data
I downloaded the file manualy and saved in my working directory, so I can just use the read.csv function to get the data

```{r}
activity <- read.csv("activity.csv",sep = ",",header = TRUE)
head(activity)
```

I use the head data just to show the first 6 lines of the data.

## Step 2: Transform the data
I would like to use the tbl_df function to transform my data, so that I can use dplyr and tidyr packages.These packages makes clean and summarizing data alot easier.

```{r,message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
activity <- tbl_df(activity)
activity
```

I use the message = false so that the packages loaded messages are not displayed in the markdown file

Next I will remove all rows with NA values, as I do not want to use those observations.

```{r}
na <- is.na(activity[["steps"]])
activity <- activity[!na,]
sum(is.na(activity))
```

I display the sum of the NA values in the activity table, the value is 0 because all the NA values was removed.

## Step 3: What is mean total number of steps taken per day?
### Calculate total number of steps taken per day

Next I will creat a table which summarizes the activity table per day using ddply

```{r}
restruc <- group_by(activity,date)
byday <- summarize(restruc, steps=sum(steps))
byday
```

I create a new table that summarizes the sum of the steps per day.

### Generate a histogram of total number of steps taken per day

```{r}
with(byday,hist(steps))
```

Histogram of total number of steps per day displayed above

### Mean and median of total number of steps per day

```{r}
with(byday,summary(steps))
```

I use the summarise function just to display a bit more information

## What is the activity pattern?

First I will summarize the number of steps by interval

```{r}
restuc2 <- group_by(activity,interval)
byinterval <- summarize(restuc2,steps=mean(steps))
byinterval
```

Next I will plot the mean-xaxis and interval-yaxis as a type 1 time series plot

```{r}
with(byinterval,plot(interval,steps,type = "l",main = "average number of steps taken, averaged across all days"))
```

Next I wil display the max interval along with the average steps

```{r}
byinterval <- arrange(byinterval,steps)
tail(byinterval,1)
```

## Imputing missing values

First I will get the original data set and report the amount of NA values

```{r}
activity2 <- read.csv("activity.csv",sep = ",",header = TRUE)
sum(is.na(activity2))
```

Next I will fill in the missing values by replacing the NA with the mean, rounded to nearest whole number, of the day it took place on

```{r}
for(i in 1:nrow(activity2)){
  test = activity2[i,1]
  if(is.na(test)){
    j = match(activity2[i,3],byinterval$interval)
    activity2[i,1]=byinterval$steps[j]
}
}
head(activity2)
sum(is.na(activity2))
```

As seen in the 6 rows of activity2 all NA has been removed and if that is not proof enough the sum of NA values in activity2 is also shown as 0

I will use the same method as used above to compute the sum of total steps per day

```{r}
restruc3 <- group_by(activity2,date)
byday2 <- summarize(restruc3,steps=sum(steps))
byday2
```

Next histogram plot as well as mean and median of total steps taken per day

```{r}
with(byday2,hist(steps))
summary(byday2$steps)
summary(byday$steps)
```

As excpected the mean and median stays the same put the percentile will change due to a larger dataset

## Are there differences in activity patterns between weekdays and weekends?

First I will add a column to the activity2 column which specifies wheather a date is weekend or weekday

```{r}
activity2$dateType <-  ifelse(as.POSIXlt(activity2$date)$wday %in% c(0,6), 'weekend', 'weekday')
activity3 <- aggregate(steps ~ interval + dateType, data=activity2, mean)
ggplot(activity3, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(.~ dateType ) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```
